Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-05-02 overleaf/compiler.tex
Cut more.

\figref{fig:compiler_framework} presents an overview of our compiler.
We start the section with a brief description of the top-level syntax and semantic restrictions in~\secref{sec:syntax}.
We proceed to describe the front end of the compiler:~\secref{sec:imp_to_ssa}
describes translation from IMP Source to SSA, and~\secref{sec:ssa_to_MPC_Source}
describes translation form SSA into MPC Source.

We describe analysis on MPC Source and backend-independent vectorization in~\secref{sec:vectorization}
and translation into MOTION code in~\secref{sec:backend}. Full details on all parts appear in~\cite{Anon_TR}.
%\ana{Need to state this more precisely.}

@@

%\begin{enumerate}
%  \item
        \paragraph{Parsing:}
        Use Python's \texttt{ast} module to parse the input source code to a Python AST.
%  \item
        \paragraph{Syntax checking:}
        Ensure that the AST matches the restricted subset defined in Section~\secref{sec:syntax}.
        %This step outputs an instance of the \texttt{restricted\_ast.Function} class, which represents our restricted subset of the Python AST.
%  \item
        \paragraph{3-address CFG conversion:}
        %\benjamin{TODO: Is this a good amount of detail?} \ana{This is good amount of detail for the Tech report. In the actual paper we'll shorten a lot and add pointers to the Tech report.}

        Convert the restricted-syntax AST to a three-address control-flow graph. The step processes for-loops, if-statements and assignments as restricted by the syntax.
\begin{comment}
        To do this, first,
        add an empty basic block to the CFG and mark it as current.
        Next, for each statement in the restricted AST's function body,
        process the statement.
        Statements can either be for-loops, if-statements, or assignments (as in~\secref{sec:syntax}).
        Rules for processing each kind of statement are given below:
        \begin{enumerate}
          \item
                \textbf{For-loops}:
                Create new basic blocks for
                the loop condition
                (the \emph{condition-block}),
                the loop body
                (the \emph{body-block}),
                and the code after the loop
                (the \emph{after-block}).
                Insert a jump from the end of the current block to the condition-block.
                Then, mark the condition-block as the current block.
                Insert a for-instruction at the end of the current block with the loop counter variable and bounds from the AST.
                Next, add an edge from the current block to the after-block labeled ``FALSE'' and
                an edge from the current block to the
                body-block labeled ``TRUE''.
                Then, set the body-block to be the current block
                and process all statements in the AST's loop body.
                Finally, insert a jump to the condition-block and set the after-block as current.
          \item
                \textbf{If-statements}:
                Create new basic blocks for
                the ``then'' statements of the if-statement
                (the \emph{then-block}),
                the ``else'' statements of the if-statement
                (the \emph{else-block}),
                and the code after the if-statement
                (the \emph{after-block}).
                At the end of the current block,
                insert a conditional jump to the then-block or else-block
                depending on the if-statement condition in the AST.
                Next,
                mark the then-block as current,
                process all then-statements,
                and add a jump to the after-block.
                Similarly,
                mark the else-block as current,
                process all else-statements,
                and add a jump to the after-block.
                Finally, set the after-block to be the current block,
                and give it a \emph{merge condition} property equal to the condition of the if-statement.
          \item
                \textbf{Assignments}:
                In the restricted-syntax AST,
                the left-hand side of assignments
                can be a variable or an array subscript.
                If it is an array subscript, e.g., {\sf A[i] = x},
                change the statement to {\sf A = Update(A, i, x)}.
                If the statement is not already three-address code,
                for each sub-expression in the right-hand side of the assignment,
                insert an assignment to a temporary variable.
        \end{enumerate}
\end{comment}        
  %\item
        \paragraph{SSA conversion:}
        Convert 3-address CFG to SSA with Cytron's algorithm~\cite{Cytron:1991}.
%\end{enumerate}

@@

Once the compiler converts the code to SSA,
it transforms $\phi$-nodes that correspond to if-statements into MUX nodes.
From the 3-address CFG conversion step,
$\phi$-nodes corresponding to if-statements will be in a basic block
with the merge condition property.
For example, if {\sf X!3 = $\phi$(X!1,X!2)} is in a block with merge condition {\sf C},
the compiler transforms it into {\sf X!3 = MUX(C, X!1, X!2)}.
Next, the compiler runs the dead code elimination algorithm from Cytron's SSA paper.

Next, the control-flow graph is \emph{linearized} into MPC Source,
which has loops but no if-then-else-statements.
This means that both branches of all if-statements are executed,
and the MUX nodes determine whether to use results from the then-block or from the else-block.
The compiler linearizes the control-flow graph with a variation of breadth-first search.
Blocks with the ``merge condition'' property are only considered
the second time they are visited,
since that will be after both branches of the if-statement are visited.
(The Python AST naturally gives rise to a translation where each conditional has exactly two targets,
and each ``merge condition'' block has exactly two incoming edges, a TRUE and a FALSE edge.
Thus, each $\phi$-node has exactly two multiplexer arguments, which dovetails into MUX.
This is in contrast with Cytron's algorithm which operates at the level of the CFG and allows for
$\phi$-nodes with multiple arguments.)
Each time the compiler visits a block,
it adds the block's statements to the MPC source.
If the block ends in a for-instruction,
the compiler recursively converts the body and code after the loop to MPC source
and adds the for-loop and code after the loop to the main MPC source.
If the block does not end in a for-instruction,
the compiler recursively converts all successor branches to MPC source and
appends to the main MPC source.

>>>>

\figref{fig:compiler_framework} presents an overview of our compiler.
%We start the section with a brief description of the top-level syntax and semantic restrictions in~\secref{sec:syntax}.
%We proceed to describe the front end of the compiler:~\secref{sec:imp_to_ssa}
%describes translation from IMP Source to SSA, and~\secref{sec:ssa_to_MPC_Source}
%describes translation form SSA into MPC Source.
This section outlines the front end, the following section describes analysis on MPC Source and backend vectorization 
and~\secref{sec:backend} outlines the back end. 
%We describe analysis on MPC Source and backend-independent vectorization in~\secref{sec:vectorization}
%and translation into MOTION code in~\secref{sec:backend}. 
Details on all parts appear in~\cite{Anon_TR}.
%\ana{Need to state this more precisely.}

@@

%\begin{enumerate}
%  \item
        \squeeze
        \paragraph{Parsing:}
        Use Python's \texttt{ast} module to parse the input source code to a Python AST.
%  \item
        \squeeze
        \paragraph{Syntax checking:}
        Ensure that the AST matches the restricted subset defined in Section~\secref{sec:syntax}.
        %This step outputs an instance of the \texttt{restricted\_ast.Function} class, which represents our restricted subset of the Python AST.
%  \item
        \squeeze
        \paragraph{3-address CFG conversion:}
        %\benjamin{TODO: Is this a good amount of detail?} \ana{This is good amount of detail for the Tech report. In the actual paper we'll shorten a lot and add pointers to the Tech report.}
        Convert the restricted-syntax AST to a three-address control-flow graph (CFG). The step processes for-loops, if-statements and assignments as restricted by the syntax.
\begin{comment}
        To do this, first,
        add an empty basic block to the CFG and mark it as current.
        Next, for each statement in the restricted AST's function body,
        process the statement.
        Statements can either be for-loops, if-statements, or assignments (as in~\secref{sec:syntax}).
        Rules for processing each kind of statement are given below:
        \begin{enumerate}
          \item
                \textbf{For-loops}:
                Create new basic blocks for
                the loop condition
                (the \emph{condition-block}),
                the loop body
                (the \emph{body-block}),
                and the code after the loop
                (the \emph{after-block}).
                Insert a jump from the end of the current block to the condition-block.
                Then, mark the condition-block as the current block.
                Insert a for-instruction at the end of the current block with the loop counter variable and bounds from the AST.
                Next, add an edge from the current block to the after-block labeled ``FALSE'' and
                an edge from the current block to the
                body-block labeled ``TRUE''.
                Then, set the body-block to be the current block
                and process all statements in the AST's loop body.
                Finally, insert a jump to the condition-block and set the after-block as current.
          \item
                \textbf{If-statements}:
                Create new basic blocks for
                the ``then'' statements of the if-statement
                (the \emph{then-block}),
                the ``else'' statements of the if-statement
                (the \emph{else-block}),
                and the code after the if-statement
                (the \emph{after-block}).
                At the end of the current block,
                insert a conditional jump to the then-block or else-block
                depending on the if-statement condition in the AST.
                Next,
                mark the then-block as current,
                process all then-statements,
                and add a jump to the after-block.
                Similarly,
                mark the else-block as current,
                process all else-statements,
                and add a jump to the after-block.
                Finally, set the after-block to be the current block,
                and give it a \emph{merge condition} property equal to the condition of the if-statement.
          \item
                \textbf{Assignments}:
                In the restricted-syntax AST,
                the left-hand side of assignments
                can be a variable or an array subscript.
                If it is an array subscript, e.g., {\sf A[i] = x},
                change the statement to {\sf A = Update(A, i, x)}.
                If the statement is not already three-address code,
                for each sub-expression in the right-hand side of the assignment,
                insert an assignment to a temporary variable.
\end{comment}        
  %\item
        \squeeze
        \paragraph{SSA conversion:}
        Convert 3-address CFG to SSA with Cytron's algorithm~\cite{Cytron:1991}.
%\end{enumerate}

@@

Once the compiler converts the code to SSA,
it transforms $\phi$-nodes that correspond to if-statements into MUX nodes.
From the 3-address CFG conversion step,
$\phi$-nodes corresponding to if-statements will be in a basic block
with the ``merge condition'' property.
E.g., if {\sf X!3 = $\phi$(X!1,X!2)} is in a block with merge condition {\sf C},
the compiler transforms it into {\sf X!3 = MUX(C, X!1, X!2)}.
Next, the compiler runs the dead code elimination algorithm from~\cite{Cytron:1991}. %Cytron's SSA paper.

Next, the control-flow graph is \emph{linearized} into MPC Source,
which has loops but no if-then-else-statements.
This means that both branches of all if-statements are executed,
and the MUX nodes determine whether to use results from the then-block or from the else-block.
The compiler linearizes the control-flow graph with a variation of breadth-first search.
%\begin{comment}
Blocks with the ``merge condition'' property are only considered
the second time they are visited,
since that will be after both branches of the if-statement are visited.
(The Python AST naturally gives rise to a translation where each conditional has exactly two targets,
and each ``merge condition'' block has exactly two incoming edges, a TRUE and a FALSE edge.
Thus, each $\phi$-node has exactly two multiplexer arguments, which dovetails into MUX.
This is in contrast with Cytron's algorithm which operates at the level of the CFG and allows for
$\phi$-nodes with multiple arguments.)
Each time the compiler visits a block,
it adds the block's statements to the MPC source.
If the block ends in a for-instruction,
the compiler recursively converts the body and code after the loop to MPC source
and adds the for-loop and code after the loop to the main MPC source.
If the block does not end in a for-instruction,
the compiler recursively converts all successor branches to MPC source and
appends to the main MPC source.
%\end{comment}

### 2022-05-02 overleaf/vectorization.tex
Cut more.

In this section, we briefly describe the analysis. Details, including edge cases and examples, can be found in the supplementary extended version~\cite{Anon_TR}.
\secref{sec:dependence} describes our dependence analysis and ~\secref{sec:expansion} describes scalar expansion,
which lifts scalars (and arrays) to the corresponding loop dimensionality to create opportunities for vectorization. 
\secref{sec:basic_vectorization} describes our core vectorization algorithm and~\secref{sec:correctness} argues correctness of the vectorization transformation.
\secref{sec:array_writes} extends vectorization with array writes.

@@

There are two key phases of the algorithm. Phase 1 inserts raise dimension and drop dimension 
operations according to def-uses. E.g., if there is an inner-to-outer dependence, it inserts $\mathit{raise\_dim}$,
and similarly, if there is an outer-to-inner dependence, it inserts $\mathit{drop\_dim}$. After this phase operations work
on arrays of the corresponding dimensionality and we optimistically vectorize all arrays. For example, 
lines 4 and 5 in Biometric~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(c) reshape the input arrays

@@

In the Histogram example, this removes the back edge from the update at 8 to the PHI node at 2. Clearly there does not exist $i < i'$ that will make 
$i$ at 9 and $i'$ at 6 and 7 equal. The back-edge from 9 to 4 stays because for every $j < j'$ 9 and 6 and 7 access the same location in {\sf res}. %\ana{Can't wrap my head around this...}

@@

Returning to the histogram example, the canonical (i.e., declared) dimensionality of {\sf res} is 1. Also, the phi-node {\sf res1 = PHI(res, res2)} is a target-less phi-node (the implication being that the inner for loop can be vectorized across $i$). After Phase 1, Vectorization produces the following code (statements are implicitly vectorized along $i$ and $j$). In a vectorized update statement, we can ignore the incoming array, {\sf res2} in this case. The update writes (in parallel) all locations of the 2-dimensional array, in this case it sets up each {\sf res3[i,j] = tmp3[i,j]}.

>>>>

In this section, we briefly describe the analysis. Details, including edge cases and examples, can be found in the supplementary extended version~\cite{Anon_TR}.
%\secref{sec:dependence} describes our dependence analysis and ~\secref{sec:expansion} describes scalar expansion,
%which lifts scalars (and arrays) to the corresponding loop dimensionality to create opportunities for vectorization. 
%\secref{sec:basic_vectorization} describes our core vectorization algorithm and~\secref{sec:correctness} argues correctness of the vectorization transformation.
%\secref{sec:array_writes} extends vectorization with array writes.

@@

There are two key phases of the algorithm. Phase 1 inserts raise dimension and drop dimension 
operations according to def-uses. E.g., if there is an inner-to-outer dependence, it inserts $\mathit{raise\_dim}$,
and similarly, if there is an outer-to-inner dependence, it inserts $\mathit{drop\_dim}$. After this phase operations work
on arrays of the corresponding dimensionality and we optimistically vectorize all arrays. For example, 
lines 4,5 in Biometric~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(c) reshape the input arrays.

@@

In Histogram, this removes the back edge from the update at 8 to the PHI node at 2. Clearly there does not exist $i < i'$ that makes
$i$ at 9 and $i'$ at 6 and 7 equal. The back-edge from 9 to 4 stays because for every $j < j'$ 9 and 6 and 7 access the same location in {\sf res}. %\ana{Can't wrap my head around this...}

@@

Returning to the histogram example, the canonical (i.e., declared) dimensionality of {\sf res} is 1. Also, the phi-node {\sf res1 = PHI(res, res2)} is a target-less phi-node (the implication being that the inner for loop can be vectorized across $i$). After Phase 1, Vectorization produces the following code (statements are implicitly vectorized along $i$ and $j$). In a vectorized update statement, we can ignore the incoming array, {\sf res2} in this case. The update writes (in parallel) all locations of the 2-d array, in this case it sets up each {\sf res3[i,j] = tmp3[i,j]}.
